The Unicode Cookbook \linebreak for Linguists

### 2018-03-09 book/reviews/responses_reviewer1.txt

[2] See Section 3.12 in which we suggest to use NFC, no-BOM and LF line breaks because of our experience in working with linguistic data in this format. A keen reviewer notes that specifying a convention for line endings and BOM is overly strict because most computing environments (now) transparently handle both alternatives. For example, using Python a  file can be decoded using the encoding “utf-8-sig”, which strips away the BOM (if present) and reads an input full in text mode, so that both line feed variants “LF” and “CRLF” will be stripped.

A2: "tabs and newline […] will lead to problems", thus they should be
explicitely disallowed!

*** Agreed! We have updated this to:

A2. A profile is a delimited text file with an obligatory header line. A minimal pro le must have a single column with the header “Grapheme”. For any additional columns, the name in the header must be speci ed.  e actual ordering of the columns is unimportant.  e header list must be delimited in the same way as the rest of the  le’s contents. Each record must be kept on a separate line. Separate lines with comments are not allowed. Comments that belong to speci c lines must be put in a separate column of the  le, e.g. add a column called comments.

@@

***

A3: The metadata file seems to be highly under-specified. At the very least,
there should be a list of recognized tags providing semantics for the data.
But since machine-readability is a requirement for the file anyway, a format
like JSON-LD [6], or even more appropriate, the JSON-LD dialect specified in
"Metadata Vocabulary for Tabular Data" [7]  may be a better option, because it
allows easy inclusion of Dublin Core [8] metadata - as well as a specification
of the exact CSV dialect used for the actual orthography profile.
It should be noted that JSON-LD metadata is the choice for datasets conforming
to the Cross-Linguistic Data Formats [9] standard.

*** Agreed. In practice, we haven't actually used these metadata files. We've updated this paragraph to "SHOULD" via RFC 2119 and we RECOMMEND they follow the practices suggested by the reviewer and implemented in CLDF:

A3. Metadata should be added in a separate UTF-8 text file using the JSON-LD dialect speci ed in Metadata Vocabulary for Tabular Data.3  is metadata format allows easy inclusion of Dublin Core metadata,4 which should be used to specify information about the orthographic description in the orthography pro le.5  e orthography pro le metadata should min- imally include provenance information including: (i) author, (ii) date, (iii) title of the pro le, and (iv) bibliographic data for resource(s) that illustrate the orthography described in the pro le. Crucially, the metadata should also specify (iv) a stable language identi er of the target language of the pro le using BCP 47/ISO 639-3 or Glo ocode as per the CLDF ontology.6 Further, the metadata  le should provide information about the orthogra- phy pro le’s structure and contents, including: (vi) its dialect description,7 (vii) the tokenization method (see below), and (ix) proper column descrip- tions,8 which describe how a column should be interpreted and processed (e.g. whether they should be processed as regular expressions; see below). Finally, in accordance with the Metadata Vocabulary for Tabular Data, the metadata’s  lename should consist of the orthography pro le’s  lename appended with “-metadata.json”.9

3 https://www.w3.org/TR/tabular-metadata/
4 http://dublincore.org/
5 http://w3c.github.io/csvw/metadata/#dfn-common-property
6 http://cldf.clld.org/v1.0/terms.rdf
7 http://w3c.github.io/csvw/metadata/#dfn-dialect-descriptions
8 http://w3c.github.io/csvw/metadata/#dfn-datatype-description
9 JSON-LD metadata is also the choice for datasets conforming to the Cross-Linguistic Data
Formats standard, see: http://cldf.clld.org/.

@@

A7: "If regular expressions are used, then […]": The specification does not
state how a processing application can determine if that is the case - other
than through additional user input. So specifiying whether regular expressions
are used should be required in the metadata file. In general, allowing regular
expressions as input is somewhat problematic, because regular expression
syntax as well as functionality is somewhat under-specified across regex
implementations in different computing environments. The closest to a standard
in this area might be "POSIX Basic Regular Expressions" [10], but these do not
include lookahead and lookbehind functionality as requred by B3. So allowing
regular expressions as part of orthography profiles may lead to different
results of applying the same profile to the same data on different platforms -
which clearly goes against the idea of this specification.

*** W

Overall, it seems somewhat strange to sacrifice
CSV standards conformance (no quoting of cell values, "LF" line endings, 
tab as separator) in order to achieve better read-/writeability by humans. This
puts a bigger burden on the authors in terms of specification (they can no 
longer just point to FC 4180) and it makes it harder to use tools like 
LibreOffice Calc or MS Excel to edit Orthography profiles.

*** Agreed (SM). The formal specification now requires a delimited text file, in which proper column descriptions are defined in the metadata JSON-LD file following:

http://w3c.github.io/csvw/metadata/#dfn-dialect-descriptions
http://w3c.github.io/csvw/metadata/#dfn-datatype-description

@@

B1: As noted for A7, to make the basic handling possible in an automated way,
there must be a way to figure out that the profile is "basic".

*** See directly above.

B2: If - as stated - "ordering […] might be crucial", it must be specified, i.e.
members a class specified as context must be tried in order of occurrence in the
profile.

*** Updated to:

Note that in some cases the ordering inside this regular expression might be crucial, therefore it must be specified as so in the metadata file.

@@

B4: I would rather spin the functionality to suggest an optimal ordering of rules
off to a separate tool, and limit the specification to apply rules in order of
occurrence. This will lead to simpler implementation and more transparent 
relation of input and output.

*** We make this clearer:

The regular expressions will be applied in the order as specified
in the profile, from top to bottom.} A software implementation can offer
help in figuring out the optimal ordering of the regular expressions, but
then it should be made explicit in the orthograpy profile because regular 
expressions are executed in order from top to bottom.

@@

Language like

    The actual implementation of the profile on some text-string will function 
    as follows: (page 83)

sounds a bit weird in a technical specification. In general, the presentation
may benefit by adopting the conventions as used for describing algorithms, i.e.
specifying
- input
- output
- a series of steps to get from input to output

*** We've updated the text to be more specification-like, as suggested above.

### 2018-03-09 book/reviews/responses_reviewer2.txt

Section Implementation
- can orthography formats handle negative lookahead/lookbehind? Note
that this is different from a positive look for the set complement if
the string looked for is longer than 1.

*** They can handle anything that can be encoded with regular expressions (and if the regular expression has that functionality).

B6.
I feel that a language which has both <ch> and <sch> will get you
different results depending on global or linear algorithm.
"Ascha" should give you "A sch a" with the linear approach, but "a s ch
a" with the global approach iff "ch" is matched before "sch".

*** This is true, thus we write:

Theoretically, these approaches will lead to different results, though in practice of actual natural language orthographies they almost always lead to the same result.

@@

Chapter 6
- "Each section has a number of subsections that ... "

*** Fixed.

- the example "aäaaöaüaa" will confuse a lot of ordinary working
linguists. There are too many vowels, and the representation of "aa" by
<x> is weird. I suggest using the German word "Schächtelchen", where you
have both <sch> and <ch> to show the importance of ordering, and a
dieresis to show decomposition (which seem to be the main didactic
aspects of this passage to me).

*** Schächtelchen is a great word and we appreciate the suggestion that "aäaaöaüaa" is confusing. So we have updated the example with "Schächtelchen" (also online).

We also built a different example:

@@

"ĉháɾã̌ctʼɛ↗ʐː| k͡p"

which captures a hypothetical form with all the intricacies of how IPA is encoded in Unicode.

@@

Section Using an orthography profile skeleton
- " It is also possible ...It is also possible": rephrase

*** Fixed.

- maybe this passage should be written with Italian or Spanish
orthography as an example since English orthography is such a mess. Why
are "ie" and "oo" digraphs? If so, would "ti" in "nation" count as a
digraph as well? And why is the second "e" in "scheme" listed at all? It
does not correspond to any phoneme (but arguably it influences the first
"e"). Also, for me, "sch" in "scheme" consist of two phonemes (/sk/),
just like "sch" in "mischief (/stS/)". But the passage "Now mischief is
parsed correctly, but scheme is wrong" suggests that the author sees
this differently???

*** Recall we are tokenizing orthographic conventions, not their phonetic or phonemic realizations.

@@

- Footnote 3: delete. Feeding and bleeding are needlessly complicating
the exposition

*** We don't agree. Feeding and bleeding is an essential concept and this is the minimum we should do to make the reader aware of it.

@@

- "It is important to realize ... ": delete. The need for escape
sequences has been discussed earlier.

*** Fixed.

- Table 6.11: I was wondering whether graphemes can be in several
classes. "Y" could be in the consonant class and in the vowel class for
instance. Can orthography profiles handle this, and if they can, how
would this be entered?

*** This would be a orthography-specific case, for which we should create an example online.

### 2018-03-09 book/chapters/implementation.tex
ready for review two

\noindent An orthography profile is tab-delimited UTF-8 text file. The first column must be labelled \texttt{Grapheme}. Each row in the \texttt{Grapheme} column specifies graphemes that may be found in the orthography of the input text. In this example, we provide additional columns IPA and XSAMPA, which are mappings from our graphemes to their IPA and XSAMPA transliterations. The final column \texttt{COMMENT} is for comments; if you want to use a tab ''quote that     string''!

@@

\noindent We set the default as the Unicode replacement character \uni{fffd}, \textsc{replacement character}.\footnote{\url{http://www.fileformat.info/info/unicode/char/fffd/index.htm}} But you can simply change this by specifying the replacement character when you load the orthography profile with the tokenizer.

@@

\noindent We can create an initial orthography profile of the German text by passing it to the \textsc{segments profile} command. The initial profile tokenizes the text on Unicode grapheme clusters, lists the frequency of each grapheme, and provides an initial mapping column by default.

@@

\noindent Next, we know a bit about German orthography and which characters combine to form German graphemes. We can use the information from our initial orthography profile to hand-curate a more precise German orthography profile that takes into account capitalization (German orthography obligatorily capitalizes nouns) and grapheme clusters, such as <sch> and <ch>. We can use the initial orthography profile above as a starting point (note the in large text the frequency column may signal errors in the input, such as typos, if they occur with a very low frequency). The initial orthography profile can be edited with a text editor or spreadsheet program. As per the orthography profile specifications (see Chapter \ref{orthography-profiles}), we can adjust rows in the \textsc{Grapheme} column and then add additional columns for translitations or comments.

@@

\noindent Using the command line \textsc{segments} function and passing it our orthography profile, we can now segment our German text example into graphemes.

@@

\noindent By providing \textsc{segments} a column for transliteration, we can convert the text into IPA.

@@

The bash executables are little files providing an interface to the R code that
can be used in a shell on a UNIX-alike machine. The location of these
executables is a bit hidden away by the install procedure of R packages. The
location can be found by the following command in R. These executables can be 
used from here, or they can be linked and/or copied to any location as wanted.

@@

For example, a good way to use the executables in a terminal is to
make softlinks (using \texttt{ln}) from the executables to a directory in your
PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
\texttt{tokenize} and \texttt{writeprofile}, and the links can be made directly 
by using Rscript to get the paths to the executables within the terminal.

@@

@@

A common workflow to use these functions is to first make a skeleton for an
orthography profile and then edit this profile by hand. For example, Table
\ref{tab:profile_skeleton1} shows the profile skeleton after a few graphemes have
been added to the file. Note that in this example, the profile is written to the
desktop, and this file has to be edited manually. We simply add a few
multigraphs to the column \texttt{Grapheme} and leave the other columns empty.
these new graphemes are then included in the graphemic parsing.

@@

There are various additional options for rule ordering implemented. Please check
the help description in R \texttt{help(tokenize)} for more details on the
possible rule ordering specifications. In summary, there are four different 
ordering options, that can also be combined:

@@

It is important to realize that different ordering of the rules does not have
so-called feeding and bleeding effects as known from finite-state rewrite
rules.\footnote{Bleeding is the effect that the application of a rule changes
the string, so as to prevent a next rule to apply. Feeding is the opposite: a
specific rule will only be applied because a previous rule changed the string
already. The interaction of rules with such feeding and bleeding effects is
extremely difficult to predict.} The graphemic parsing advocated here works
crucially different from rewrite rules in that there is nothing being rewritten:
each line in an orthography profile specifies a grapheme to be captured in the 
string. All lines in the profile are processed in a specified order (as determined
by the option \texttt{ordering}). At the processing of a specific line, all 
matching graphemes in the data are marked as captured, but not changed. 
Captured parts cannot be captured again, but they can still be used to match 
contexts of other lines in the profile. Only when all lines are processed the 
captured graphemes are separated (and possibly transliterated). In this way the 
result of the application of the rules' is rather easy to predict.

@@

   \item \textsc{result\_strings.tsv}:\\ A tab-separated file with the original
         and the tokenized/transliterated strings

@@

   \item \textsc{result\_errors.tsv}:\\ A tab-separated file with all original
         strings that contain unmatched parts. Unmatched parts are indicated
         with the character as specified with the option \texttt{missing}. By
         default the character \textsc{double question mark} <⁇> at
         \uni{2047} is used. When there are no errors, then this file is 
         absent.

@@

It is important to realize that with the option \texttt{regex = TRUE} all
content in the profile is treated as regular expressions, so the characters with
special meaning in regular expressions should be either omitted or escaped (by
putting a <\ \backslash\ > \textsc{reverse solidus} at \uni{005C} before the
character). Specifically, this concerns the following characters:

@@

When it is expected that context might be important for a profile, then the
profile skeleton can be created with columns prepared for the contextual
specifications. This is done by using the option \texttt{editing = TRUE} (cf.
Table~\ref{tab:profile_editing_1} for a toy profile of some Italian words).

@@

Given a set of graphemes, there are at least two different methods to tokenize
strings. The first is called \texttt{method = "global"}. This approach
takes the first grapheme in the profile, then matches this grapheme globally at
all places in the string, and then turns to the next string in the profile. To
us, this seems to be the most intuitive way for linguists to deal with graphemic
parsing. The other approach is called \texttt{method = "linear"}. This
approach walks through the string from left to right. At the first character it
looks through all graphemes whether there is any match, and then walks further
to the end of the match and starts again. This approach is more alike to
finite-state rewrite rules (though note that it still works differently from
such rewrite rules, see Section~\ref{rule-ordering}). The global method is used 
by default.

@@

The problem can be exemplified by a sequence <bbbb> in which a grapheme <bb>
should be matched. With the default \texttt{regex = FALSE} there are three
possible matches, but with \texttt{regex = TRUE} only the first two <b>s or the
last two <b>s are matched. The middle two <b>s are not matched because they
overlap with the other matches. In the example below this leads to an error,
because the second <bb> is not matched. However, we have not been able to
produce a real example in any natural language in which this limitation might
lead to an error.

>>>>

\noindent An orthography profile is delimited UTF-8 text file (here we use tab as a delimiter for reading ease). The first column must be labelled \texttt{Grapheme}. Each row in the \texttt{Grapheme} column specifies graphemes that may be found in the orthography of the input text. In this example, we provide additional columns IPA and XSAMPA, which are mappings from our graphemes to their IPA and XSAMPA transliterations. The final column \texttt{COMMENT} is for comments; if you want to use a tab ''quote that     string''!

@@

\noindent We set the default as the Unicode replacement character \uni{fffd}, \texttt{replacement character}.\footnote{\url{http://www.fileformat.info/info/unicode/char/fffd/index.htm}} But you can simply change this by specifying the replacement character when you load the orthography profile with the tokenizer.

@@

\noindent We can create an initial orthography profile of the German text by passing it to the \texttt{segments profile} command. The initial profile tokenizes the text on Unicode grapheme clusters, lists the frequency of each grapheme, and provides an initial mapping column by default.

@@

\noindent Next, we know a bit about German orthography and which characters combine to form German graphemes. We can use the information from our initial orthography profile to hand-curate a more precise German orthography profile that takes into account capitalization (German orthography obligatorily capitalizes nouns) and grapheme clusters, such as <sch> and <ch>. We can use the initial orthography profile above as a starting point (note the in large text the frequency column may signal errors in the input, such as typos, if they occur with a very low frequency). The initial orthography profile can be edited with a text editor or spreadsheet program. As per the orthography profile specifications (see Chapter \ref{orthography-profiles}), we can adjust rows in the \texttt{Grapheme} column and then add additional columns for translitations or comments.

@@

\noindent Using the command line \texttt{segments} function and passing it our orthography profile, we can now segment our German text example into graphemes.

@@

\noindent By providing \texttt{segments} a column for transliteration, we can convert the text into IPA.

@@

The bash executables are little files providing an interface to the R code that
can be used in a shell on a UNIX-alike machine. The location of these
executables is a bit hidden away by the install procedure of R packages. The
location can be found by the following command in R.

@@

These executables can be 
used from here, or they can be linked and/or copied to any location as wanted. 
For example, a good way to use the executables in a terminal is to
make softlinks (using \texttt{ln}) from the executables to a directory in your
PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
\texttt{tokenize} and \texttt{writeprofile}, and the links can be made directly 
by using Rscript to get the paths to the executables within the terminal.

@@

To make the functionality even more accessible, we have prepared webapps with 
the \texttt{shiny} framework for the R functions. The webapps are 
included inside the \texttt{qlcData} package and can be started with the 
helper function (in R): \texttt{launch\_shiny('tokenize')}.

@@

A common workflow to use these functions is to first make a skeleton for an
orthography profile and then edit this profile by hand. For example, Table
\ref{tab:profile_skeleton1} shows the profile skeleton after a few graphemes have
been added to the file. Note that in this example, the profile is written to the
desktop, and this file has to be edited manually. We simply add a few
multigraphs to the column \texttt{Grapheme} and leave the other columns empty.
These new graphemes are then included in the graphemic parsing.

@@

There are various additional options for rule ordering implemented. Please check
the help description in R, i.e.\ \texttt{help(tokenize)}, for more details on the
possible rule ordering specifications. In summary, there are four different 
ordering options, that can also be combined:

@@

Note that different ordering of the rules does not have
so-called feeding and bleeding effects as known from finite-state rewrite
rules.\footnote{Bleeding is the effect that the application of a rule changes
the string, so as to prevent a next rule to apply. Feeding is the opposite: a
specific rule will only be applied because a previous rule changed the string
already. The interaction of rules with such feeding and bleeding effects is
extremely difficult to predict.} The graphemic parsing advocated here works
crucially different from rewrite rules in that there is nothing being rewritten:
each line in an orthography profile specifies a grapheme to be captured in the 
string. All lines in the profile are processed in a specified order (as determined
by the option \texttt{ordering}). At the processing of a specific line, all 
matching graphemes in the data are marked as captured, but not changed. 
Captured parts cannot be captured again, but they can still be used to match 
contexts of other lines in the profile. Only when all lines are processed the 
captured graphemes are separated (and possibly transliterated). In this way the 
result of the application of the rules is rather easy to predict.

@@

   \item \textsc{result\_strings.tsv}:\\ A tab-separated file with the original
         and the tokenized/transliterated strings.

@@

   \item \textsc{result\_errors.tsv}:\\ A tab-separated file with all original
         strings that contain unmatched parts. Unmatched parts are indicated
         with the character as specified with the option \texttt{missing}. By
         default the character \textsc{double question mark} <⁇> at
         \uni{2047} is used. When there are no errors, this file is 
         absent.

@@

Note that with the option \texttt{regex = TRUE} all
content in the profile is treated as regular expressions, so the characters with
special meaning in regular expressions should be either omitted or escaped (by
putting a <\ \backslash\ > \textsc{reverse solidus} at \uni{005C} before the
character). Specifically, this concerns the following characters:

@@

When it is expected that context might be important for a profile, then the
profile skeleton can be created with columns prepared for the contextual
specifications. This is done by using the option \texttt{editing = TRUE}~(cf.\
Table~\ref{tab:profile_editing_1} for a toy profile of some Italian words).

@@

Given a set of graphemes, there are at least two different methods to tokenize
strings. The first is called \texttt{method = "global"}. This approach
takes the first grapheme in the profile, then matches this grapheme globally at
all places in the string, and then turns to the next string in the profile. The 
other approach is called \texttt{method = "linear"}. This
approach walks through the string from left to right. At the first character it
looks through all graphemes whether there is any match, and then walks further
to the end of the match and starts again. This approach is more akin to
finite-state rewrite rules (though note that it still works differently from
such rewrite rules, see above). The global method is used 
by default in the R implementation.

@@

The problem can be exemplified by a sequence <bbbb> in which a grapheme <bb>
should be matched. With the default \texttt{regex = FALSE} there are three
possible matches, but with \texttt{regex = TRUE} only the first two <b>'s or the
last two <b>'s are matched. The middle two <b>'s are not matched because they
overlap with the other matches. In the example below this leads to an error,
because the second <bb> is not matched. However, we have not been able to
produce a real example in any natural language in which this limitation might
lead to an error.
